---
title: "Creating a Custom Function to Calculate Indices"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{custom-indices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(eemanalyzeR)
```

## Introduction

To maximum flexibility in calculating indices, we've enabled the [get_indices](../html/get-indices.html) function to accept custom written functions, allowing users to create their own function and calculate indices that may not be included in existing methods.

## Function Requirements

1.  The function needs to have the following arguments:
    a.  eemlist: an eemlist object containing EEM's data.
    b.  abslist: an abslist object containing absorbance data.
    c.  cuvle: the cuvette (path) length in cm.
2.  The function must return a list with the following items:
    a.  abs_index: the indices calculated from the absorbance data
    b.  eem_index: the indices calculated from the EEMs data

abs_index and eem_index should both be a data.frame with the following columns:

-   sample_name: a character, the name of the sample

-   meta_name: a character, the name of the sample in the metadata if metadata has been added, otherwise the sample name again

-   index: a character, the name of the index being reported

-   value: the value of the index, also accepts flag values (see details below)

## Helper Functions

The eemanalyzeR package contains a number of functions helpful in creating custom indices.

To get the absorbance values at a certain wavelength the [get_absorbance](../html/get-absorbance.html) function can be used. This function can additionally provide specific absorbance (like SUVA 254).

```{r}
#add metadata to get DOC values
abslist <- add_metadata(metadata, example_absorbance)

#absorbance at 254 nm
a254 <- eemanalyzeR::get_absorbance(abslist, 254) 
a254 

#specific absorbance at 254 nm
suva254 <- eemanalyzeR::get_absorbance(abslist, 254, suva=TRUE) 
suva254
```

A similar function exists for getting fluorescence values from EEMs data: [get_fluorescence](../html/get-fluorescence.html). Here, you can specify a range of excitation and/or emission wavelengths. If stat is set to `sum` the sum of fluorescence will be returned, otherwise if `max` is used the maximum fluorescence across that range will be returned.

```{r}
#get maximum fluorescence across ranges for Peak A and D
pA <- get_fluorescence(example_eems, ex=250:260, em=380:480, stat="max")
pD <- get_fluorescence(example_eems, ex=390, em=509, stat="max")

#get sum of fluorescence across range of Peak A
pA_sum <- get_fluorescence(example_eems, ex=250:260, em=380:480, stat="sum")
```

## QA/QC Flags

While the [get_indices](../html/get-indices.html) function includes some QA/QC checks (negative numbers, missing values, below noise, outside expected range, etc.), there are some checks that are easier to do when calculating the indices, or you may wish to include custom flags or checks. To pass flags from your index function to the output, they need to get put in the `value` column of the index data.frames.

There are a few rules to follow with flags:

1.  Flags should have the form text\_## (DOC_01, DATA_04, etc.)

2.  If an index value is NA and you want to flag with a reason why it is NA, replace the NA value with that flag value.

3.  If you wish to report a value, but still include a flag, report the value in the form: value_flag (0.042_DATA_01, 43.1_DOC_01, etc.)

### Helper Functions

There are few helper functions within the eemanalyzeR package that will provide some common flags.

First is the [get_ratios](../html/get-ratios.html) function. This can be helpful when calculating things like peak ratios, where a missing value will result in an `NA` or a value of 0 in the denominator will result in an infinite value.

```{r}
#calculate the ratio of Peak A to Peak T
 pA <- get_fluorescence(example_eems, ex=250:260, em=380:480)
 pT <- get_fluorescence(example_eems, ex=270:280, em=320:350)
 rAT <- get_ratios(pA, pT)
 rAT
 
  #if Peak T is all 0, will return DATA_03 flag, indicating index couldn't 
    #calculated due to zero in denominator
    rAT <- get_ratios(pA, 0)
    rAT
 
#calculate HIX 
  sum_high <- get_fluorescence(example_eems, ex=254, em=435:480, stat = "sum")
  sum_low <- get_fluorescence(example_eems, ex=254, em=300:345, stat="sum")
  HIX <- get_ratios(sum_high,sum_low)
  HIX
  
  #if sum_low is NA, will return DATA_01 flag indicating the index couldn't be 
    #calculated due to missing data
    HIX <- get_ratios(sum_high, NA)
    HIX

```

There's also a function that will flag indices that can't be calculated because data didn't exist for the requested wavelengths: [flag_missing](../html/flag-missing.html). This is useful to check when calculating indices to flag why certain indices might be returning `NA` values.

```{r}
#checking absorbance data
  #data exists
  flag_missing(example_absorbance, wl=400) 

 #data doesn't exist
  flag_missing(example_absorbance, wl=100)
  
 #some data exists, still calculate
  flag_missing(example_absorbance, wl=100:254, all=FALSE)

#checking fluorescence data
  #data exists
  flag_missing(example_eems, ex=270:280, em=300:320) 
  
  #data doesn't exist
  flag_missing(example_eems, ex=100:150, em=300:320) 
  
  #some data exists, still calculate
  flag_missing(example_eems, ex=100:350, em=300:320, all=FALSE) 
```

Once we've gotten the values and the flags, the data needs to be formatted in a specific way to input into the [get_indices](../html/get-indices.html) function. To do this we can use the [format_index](../html/format-index.html) function.

```{r}
  ex <- 240:260
  em <- 300:320
  vals <- get_fluorescence(example_eems, ex, em, stat = "max")
  flags <- flag_missing(example_eems, ex=ex, em=em, all=FALSE)
  index_formatted <- format_index(example_eems, "test_index", vals, flags)
```

## Examples

Alright, now that we've reviewed all the helper functions and the requirements for the custom function let's do an example. Let's create a custom function to calculate the maximum fluorescence for the five regions defined in Chen et al. 2003:

```{r echo=FALSE, include=TRUE}
indices <- data.frame(region=c("I", "II", "III", "IV", "V"), 
                      excitation=c("<250", "<250", "<250", ">250", ">250"),
                      emission=c("<300","330-380", ">380", "<380",">380"))

colnames(indices) <- c("Region", "Excitation (nm)", "Emission (nm)")
knitr::kable(indices)
```

First let's turn the different regions into a list of excitation and emission wavelengths.

```{r}
chen_regions <- list(r1 = list(ex=240:250, em=245:300),
                     r2 = list(ex=240:250, em=330:380),
                     r3 = list(ex=240:250, em=380:550),
                     r4 = list(ex=250:800, em=245:380),
                     r5 = list(ex=250:800, em=380:550))
```

We can use the [get_fluorescence](../html/get-fluorescence.html) helper function to calculate the maximum fluorescence in these regions. For a single region that would look like:

```{r}
get_fluorescence(example_eems, chen_regions[[1]]$ex, chen_regions[[1]]$em, stat="max")

```
To get all the indices let's use the lapply function. 

```{r}
results <- lapply(names(chen_regions), function(region_name) {
  region <- chen_regions[[region_name]]
  get_fluorescence(example_eems, region$ex, region$em, stat = "max")
  })

```

While we're calculating each index, let's also use the [flag_missing](../html/flag-missing.html) function to see if there's any data missing. Since we're finding the maximum value, we'll set `all` to `FALSE` so it will report a value if there's some data. Lastly, we'll use the [format_index](../html/format-index.html) function to make sure everything is correctly formatted.

```{r}
results <- lapply(names(chen_regions), function(region_name){
  #get values
  region <- chen_regions[[region_name]]
  vals <- get_fluorescence(example_eems, region$ex, region$em, stat = "max")

  #get flags
  flags <- flag_missing(example_eems, ex=region$ex, em=region$em, all=FALSE)

  #add sample names and make into data.frame (get index name)
  res <- format_index(example_eems, region_name, vals, flags)
    
  #return res
  return(res)
  })

results <- do.call(rbind, results)
```


Lastly, we just need to return the results as a list. Since we don't have any absorbance data we'll just make it NA.
```{r}
  list(abs_index = NA,
       eem_index=results)
```

Great, that all looks good, now we just need to combine all that code to a custom function. 
```{r}
chen2003 <- function(eemlist, abslist, cuvle=1){
  #making sure eemlist and abslist are correct objects
    stopifnot(.is_eemlist(eemlist), .is_abslist(abslist), is.numeric(cuvle), all(sapply(eemlist, attr, "is_doc_normalized"))==FALSE)
  #specify regions
    chen_regions <- list(r1 = list(ex=240:250, em=245:300),
                     r2 = list(ex=240:250, em=330:380),
                     r3 = list(ex=240:250, em=380:550),
                     r4 = list(ex=250:800, em=245:380),
                     r5 = list(ex=250:800, em=380:550))
  
  #get results as a data.frame
    results <- lapply(names(chen_regions), function(region_name){
    #get values
    region <- chen_regions[[region_name]]
    vals <- get_fluorescence(example_eems, region$ex, region$em, stat = "max")
  
    #get flags
    flags <- flag_missing(example_eems, ex=region$ex, em=region$em, all=FALSE)
  
    #add sample names and make into data.frame (get index name)
    res <- format_index(example_eems, region_name, vals, flags)
      
    #return res
    return(res)
    })
  
  #combine lists together
    results <- do.call(rbind, results)

  return(list(abs_index = NA,
         eem_index=results))
}
```

Now let's test our custom function with the [get_indices](../html/get-indices.html) function. 
```{r}
indices <- get_indices(example_eems, example_absorbance, index_method = chen2003, return="wide")

indices$eem_index
```

## References

Chen, W., Westerhoff, P., Leenheer, J. A., & Booksh, K. (2003). Fluorescence excitation-emission matrix regional integration to quantify spectra for dissolved organic matter. Environmental Science & Technology, 37(24), 5701–5710. <https://doi.org/10.1021/es034354c>
