---
title: "Calculating Indices with eemanalyzeR Method"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{eemanalyzeR-indices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(include = FALSE)
library(eemanalyzeR)
```

This vignette provides detailed documentation for the optical indices generated by
the eemanalyzeR function 'get_indices' using the default index method 'eemanalyzeR'. 

## What are optical indices? 

Excitation emission matrices (EEMs) and absorbance spectra can be large, unwieldy datasets which can be challenging to interpret. While procedures like parallel factor analysis (PARAFAC) can be used (Stedmon et al. 2003), these analyses can be complex, time consuming, and require substantial computing power. An alternative method of interpreting optical data is to use absorbance and fluorescence indices which are a single value calculated from the absorbance or fluorescence spectra. The benefit is a single value which is easily derived and plotted. However, interpretation can be challenging as many of these metrics were derived in specific environments, and the interpretation may not be transferable to other environments or sample types (Gabor et al. 2014, Serène et al. 2025). 

## Absorbance Indices 

```{r echo=FALSE, include=TRUE}
abs_index <- data.frame('Index Name'=c("SUVA 254", "SUVA 280", "SVA 412",
                               "Spectral Slope (275-295)",
                               "Spectral Slope (350-400)",
                               "Spectral Slope Ratio", 
                               "E~2~/E~3~", "E~4~/E~6~"
                               ), 
                        'Index Abbreviation' =c("SUVA254", "SUVA280", "SVA412", "S275_295",
                                    "S350_400", "SR", "E2_E3", "E4_E6"),
                        'Interpretation'=c("Proxy for aromaticity", 
                                    "Proxy for aromaticity",
                                    "Proxy for aromaticity",
                                    "Related to molecular weight and aromaticity",
                                    "Related to molecular weight and aromaticity", 
                                    "Related to molecular weight", 
                                    "Related to molecular weight and aromaticity",
                                    "Related to humic-like organic matter"
                                    ))
colnames(abs_index) <- c("Index Name", "Index Abbreviation", "Interpretation")
knitr::kable(abs_index)
```

### SUVA 254 (SUVA254)
**Calculation:** The absorbance at 254 nm per unit of carbon, giving the index units of L mg-C^-1^ m^-1^.

**Interpretation:** SUVA 254 is considered an proxy for aromatic content of the dissolved organic matter, where higher values indicate higher aromaticity. It has also been found to be correlated with some disinfection by-product precursors (Rostad et al. 2000). 

**Source(s):** Weishaar et al. 2003


### SUVA 280 (SUVA280) 
**Calculation:** The absorbance at 280 nm per unit of carbon, giving the index units of L mg-C^-1^ m^-1^.

**Interpretation:** Similar to SUVA 254, SUVA 280 is considered an proxy for aromatic content of the dissolved organic matter, where higher values indicate higher aromaticity. This index has been proposed as an alternative to SUVA 254, with some evidence that 280 nm may be a better wavelength to determine aromaticity at since $\pi$ to $\pi*$ transitions occur here for many aromatic molecules (Chin et al. 1994).

**Source(s):** Chin et al. 1994, Hansen et al. 2016

### SVA 412 (SVA412)

**Calculation:** The absorbance at 412 nm per unit of carbon, giving the index units of L mg-C^-1^ m^-1^.

**Interpretation:** As with the other specific absorbance metrics, this can be used as a proxy for aromatic content of DOM where higher values indicate a higher amount of aromaticity. 

**Source(s):** Hansen et al. 2016


### Spectral Slope Between 275 to 295 nm (S275_295)
**Calculation:** Found by fitting a non-linear exponential function to the absorbance spectrum between 275-295 nm. 

**Interpretation:** Higher values are typically associated with lower molecular weight materials and/or lower aromaticity.

**Source(s):** Helms et al. 2008, Hansen et al. 2016

### Spectral Slope Between 350 to 400 nm (S350_400) 
**Calculation:** Found by fitting a non-linear exponential function to the absorbance spectrum between 350-400 nm

**Interpretation:** Higher values are typically associated with lower molecular weight materials and/or lower aromaticity.

**Source(s):** Helms et al. 2008, Hansen et al. 2016

### Spectral Slope Ratio (SR)
**Calculation:** The ratio of the slope between 275 to 295 nm to the slope between 350 to 400 nm.

**Interpretation:** Negatively correlated to molecular weight and generally increases on irradiation.

**Source(s):** Helms et al. 2008, Hansen et al. 2016

### E~2~/E~3~ Ratio (E2_E3)
**Calculation:** The ratio of absorbance at 250 nm to absorbance at 365 nm.

**Interpretation:** This metric has been related to molecular size and aromaticity where, as E~2~/E~3~ increases, the aromaticity and molecular weight decreases.

**Source(s):** Peuravuori and Pihlaja 1997

### E~4~/E~6~ Ratio (E4_E6)
**Calculation:** The ratio of absorbance at 465 nm to absorbance at 665 nm.

**Interpretation:** While this metric was initial used to describe the aromaticity of organic matter, it's been found that this metric is a better descriptor of the amount of humic-like organic matter.

**Source(s):** Helms et al. 2008, Chen et al. 1977 


## Fluorescence Indices 

```{r echo=FALSE, include=TRUE}
eem_index <- data.frame('Index Name'=c("Peak B", "Peak T", "Peak A", "Peak M", "Peak C",
                                       "Peak D", "Peak E", "Peak N", 
                                       "Ratio of Peak A to Peak T",
                                       "Ratio of Peak C to Peak A",
                                       "Ratio of Peak C to Peak M",
                                       "Ratio of Peak C to Peak T",
                                       "Fluorescence Index", "Humification Index-Zsolnay",
                                       "Humification Index-Ohno",
                                       "Freshness Index", "Biological Index"), 
                        'Index Abbreviation' =c("pB", "pT", "pA", "pM","pC","pD",
                                                "pE", "pN","rAT", "rCA", "rCM", "rCT", 
                                                "FI", "HIX", "HIX_ohno", "fresh", "BIX"),
                        'Interpretation'=c("tyrosine-like, protein-like organic matter",
                                           "tryptophan-like, protein-like organic matter",
                                           "UV humic-like organic matter",
                                           "marine humic-like organic matter", 
                                           "visible humic-like organic matter", 
                                           "soil fulvic acid-like organic matter",
                                          "soil fulvic acid-like organic matter",
                                          "related to phytoplankton productivity",
                                          "ratio of humic-like to fresh organic matter",
                                          "ratio of humic-like to fulvic-like organic matter",
                                          "amount of blueshifted organic matter",
                                          "related to biochemical oxygen demand",
                                          "terrestrial versus microbial sources", 
                                          "indication of humic substances",
                                          "indication of humic substances", 
                                          "indication of recently produced organic matter",
                                          "indicator of autotrophic activity"

                                    ))
colnames(eem_index) <- c("Index Name", "Index Abbreviation", "Interpretation")
knitr::kable(eem_index)
```

### Peak B (pB)

**Calculation:** Maximum fluorescence intensity between ex: 270-280 nm and em: 300-320 nm.

**Interpretation:** This peak is typically associated with tyrosine-like, protein-like
organic matter.

**Source(s):** Coble 1996, Gabor et al. 2014

### Peak T (pT)

**Calculation:** Maximum fluorescence intensity between ex: 270-280 nm and em: 320-350 nm.

**Interpretation:** This peak is typically associated with tryptophan-like, protein-like
organic matter.

**Source(s):** Coble 1996, Gabor et al. 2014

### Peak A (pA)

**Calculation:** Maximum fluorescence intensity between ex: 250-260 nm and em: 380-480 nm.

**Interpretation:** This peak is typically associated with UV humic-like
organic matter.

**Source(s):** Coble 1996, Gabor et al. 2014

### Peak M (pM)

**Calculation:** Maximum fluorescence intensity between ex: 310-320 nm and em: 380-420 nm.

**Interpretation:** This peak is typically associated with marine humic-like
organic matter. However, this component has been observed in non-humic environments. 

**Source(s):** Coble 1996, Gabor et al. 2014

### Peak C (pC)

**Calculation:** Maximum fluorescence intensity between ex: 330-350 nm and em: 420-480 nm.

**Interpretation:** This peak is typically associated with visible humic-like organic matter. 

**Source(s):** Coble 1996, Gabor et al. 2014 

### Peak D (pD)

**Calculation:** Fluorescence intensity at ex: 390 nm and em: 509 nm.

**Interpretation:** This peak is typically associated with soil fulvic acid-like organic matter. 

**Source(s):** Stedmon et al. 2003, Aiken, 2014 

### Peak E (pE)

**Calculation:** Fluorescence intensity at ex: 455 nm, em: 521 nm.

**Interpretation:** This peak is typically associated with soil fulvic acid-like organic matter. 

**Source(s):** Stedmon et al. 2003, Aiken, 2014 

### Peak N (pN)

**Calculation:**  Fluorescence intensity at ex: 280 nm, em: 370 nm.

**Interpretation:** This peak is typically associated with associated with phytoplankton productivity. 

**Source(s):** Stedmon et al. 2003, Aiken, 2014

### Ratio of Peak A to Peak T (rAT)

**Calculation:** The ratio of Peak A to Peak T.

**Interpretation:** Related to the amount of humic-like (recalcitrant) to fresh (liable) dissolved organic matter.

**Source(s):** Hansen et al. 2016 

### Ratio of Peak C to Peak A (rCA)

**Calculation:** The ratio of Peak C to Peak A.

**Interpretation:** Related to the amount of the amount of humic-like to fulvic-like dissolved organic matter.

**Source(s):** Hansen et al. 2016 

### Ratio of Peak C to Peak M (rCM)

**Calculation:** The ratio of Peak C to Peak M.

**Interpretation:** Related to the amount of diagenetically altered (blue shifted) dissolved organic matter.

**Source(s):** Hansen et al. 2016 

### Ratio of Peak C to Peak T (rCT)

**Calculation:** The ratio of Peak C to Peak T.

**Interpretation:** Related to the amount of humic-like (recalcitrant) to fresh (liable) dissolved organic matter. Can be used to identify sewage effluents and may indicate biochemical oxygen demand.

**Source(s):** Baker 2001, Hansen et al. 2016 

### Fluorescence Index (FI)

**Calculation:** The ratio of fluorescence of em: 470 nm to em: 520 nm at ex: 370 nm.

**Interpretation:** Has typically been used to identify the relative contributions of terrestrial to microbial dissolved organic matter sources, where higher values indicate more microbial sources. Typically microbially derived organic matter has values around 1.9 and terrestrial sources have values around 1.4-1.5. However, newer work suggests that this metric can also be used to assess the complexity and aromaticity of humic-like dissolved organic matter, with lower values indicating more complex and aromatic organic matter.

**Source(s):** McKnight et al. 2001, Cory and McKnight 2005, Serène et al. 2025 

### Humification Index-Zsolnay (HIX)

**Calculation:** The ratio of the sum of fluorescence between em: 435-480 nm and em: 300-345 at ex: 254 nm.

**Interpretation:** Has typically been used as an indication of humic substances or extent of humification, described by the H/C ratio where a higher value indicates a lower H/C ratio and a higher degree of humification. However, more recent work suggests this metric is related to the ratio of complex, aromatic molecules to simple, less aromatic molecules, where higher values indicate more aromatic and polycondensed organic matter.

**Source(s):** Zsolnay et al. 1999, Serène et al. 2025 

### Humification Index-Ohno (HIX_ohno)

**Calculation:** The ratio of the sum of fluorescence between em: 435-480 nm and the sum of em: 300-345 and em: 435-480 nm at ex: 254 nm.

**Interpretation:** Interpreted the same way as the other calculation for humification index. Both versions of HIX are used throughout the literature. The version proposed by Ohno is better when samples have higher absorbance because it accounts for inner filter effects more effectively.

**Source(s):** Ohno et al. 2002, Gabor et al. 2014

### Freshness Index or β/α (fresh)

**Calculation:** The ratio of the fluorescence at em: 380 nm and the sum of em: 420-435 nm at ex: 310 nm.

**Interpretation:** An indication of recently produced DOM from biological activity, where higher values indicate more recently produced DOM.

**Source(s):** Wilson and Xenopoulos 2009

### Biological Index (BIX)

**Calculation:** The ratio of the fluorescence at em: 380 nm and em: 430 nm at ex: 310 nm.

**Interpretation:** Traditionally used as an indicator of autotrophic productivity where values above 1 indicate recently produced autochthonous dissolved organic matter. However, recent work suggests that the biological index should be used as an indicator of humic origin, were higher values indicate more microbial origins while lower values indicate more lignin or soil derived organic matter.

**Source(s):** Huguet et al. 2009, Serène et al. 2025

## References 
Aiken, G. (2014). Fluorescence and Dissolved Organic Matter: A Chemist’s Perspective. In A. Baker, D. M. Reynolds, J. Lead, P. G. Coble, & R. G. M. Spencer (Eds.), Aquatic Organic Matter Fluorescence (pp. 35-74). Cambridge: Cambridge University Press. https://doi.org/10.1017/CBO9781139045452.005

<br>

Baker, A. (2001). Fluorescence Excitation−Emission Matrix Characterization of Some Sewage-Impacted Rivers. Environmental Science & Technology, 35(5), 948-953. https://doi.org/10.1021/es000177t

<br>

Chen, Y., Senesi, N., & Schnitzer, M. (1977). Information Provided on Humic Substances by E4/E6 Ratios. Soil Science Society of America Journal, 41(2), 352-358. https://doi.org/10.2136/sssaj1977.03615995004100020037x

<br>

Chin, Y.-Ping., Aiken, George., & O’Loughlin, Edward. (1994). Molecular Weight, Polydispersity, and Spectroscopic Properties of Aquatic Humic Substances. Environmental Science & Technology, 28(11), 1853-1858. https://doi.org/10.1021/es00060a015

<br>

Coble, P. G. (1996). Characterization of marine and terrestrial DOM in seawater using excitation-emission matrix spectroscopy. Marine Chemistry, 51(4), 325-346. https://doi.org/10.1016/0304-4203(95)00062-3

<br>

Cory, R. M., & McKnight, D. M. (2005). Fluorescence Spectroscopy Reveals Ubiquitous Presence of Oxidized and Reduced Quinones in Dissolved Organic Matter. Environmental Science & Technology, 39(21), 8142–8149. https://doi.org/10.1021/es0506962

<br>

Gabor, R. S., Baker, A., McKnight, D. M., & Miller, M. P. (2014). Fluorescence Indices and Their Interpretation. In Aquatic Organic Matter Fluorescence (pp. 303-338). Cambridge University Press. https://doi.org/10.1017/CBO9781139045452.015

<br>

Hansen, A. M., Kraus, T. E. C., Pellerin, B. A., Fleck, J. A., Downing, B. D., & Bergamaschi, B. A. (2016). Optical properties of dissolved organic matter (DOM): Effects of biological and photolytic degradation. Limnology and Oceanography, 61(3), 1015-1032. https://doi.org/10.1002/lno.10270

<br>

Helms, J. R., Stubbins, A., Ritchie, J. D., Minor, E. C., Kieber, D. J., & Mopper, K. (2008). Absorption spectral slopes and slope ratios as indicators of molecular weight, source, and photobleaching of chromophoric dissolved organic matter. Limnology and Oceanography, 53(3), 955-969. https://doi.org/10.4319/lo.2008.53.3.0955

<br>

Huguet, A., Vacher, L., Relexans, S., Saubusse, S., Froidefond, J. M., & Parlanti, E. (2009). Properties of fluorescent dissolved organic matter in the Gironde Estuary. Organic Geochemistry, 40(6), 706-719. https://doi.org/10.1016/j.orggeochem.2009.03.002

<br>

McKnight, D. M., Boyer, E. W., Westerhoff, P. K., Doran, P. T., Kulbe, T., & Andersen, D. T. (2001). Spectrofluorometric characterization of dissolved organic matter for indication of precursor organic material and aromaticity. Limnology and Oceanography, 46(1), 38-48. https://doi.org/10.4319/lo.2001.46.1.0038

<br>

Ohno, T. (2002). Fluorescence inner-filtering correction for determining the humification index of dissolved organic matter. Environmental Science & Technology, 36(4), 742-746. https://doi.org/10.1021/es0155276

<br>

Peuravuori, J., & Pihlaja, K. (1997). Molecular size distribution and spectroscopic properties of aquatic humic substances. Analytica Chimica Acta, 337(2), 133-149. https://doi.org/10.1016/S0003-2670(96)00412-6

<br>

Serène, L., Mazzilli, N., Batiot-Guilhe, C., Emblanch, C., Gillon, M., Babic, M., et al. (2025). Questioning calculation and interpretation of fluorescence indices in natural waters studies. Journal of Hydrology, 650, 132524. https://doi.org/10.1016/j.jhydrol.2024.132524

<br>

Stedmon, C. A., Markager, S., & Bro, R. (2003). Tracing dissolved organic matter in aquatic environments using a new approach to fluorescence spectroscopy. Marine Chemistry, 82(3), 239-254. https://doi.org/10.1016/S0304-4203(03)00072-0

<br>

Weishaar, J. L., Aiken, G. R., Bergamaschi, B. A., Fram, M. S., Fujii, R., & Mopper, K. (2003). Evaluation of Specific Ultraviolet Absorbance as an Indicator of the Chemical Composition and Reactivity of Dissolved Organic Carbon. Environmental Science & Technology, 37(20), 4702-4708. https://doi.org/10.1021/es030360x

<br>

Wilson, H. F., & Xenopoulos, M. A. (2009). Effects of agricultural land use on the composition of fluvial dissolved organic matter. Nature Geoscience, 2(1), 37-41. https://doi.org/10.1038/ngeo391

<br>

Zsolnay, A., Baigar, E., Jimenez, M., Steinweg, B., & Saccomandi, F. (1999). Differentiating with fluorescence spectroscopy the sources of dissolved organic matter in soils subjected to drying. Chemosphere, 38(1), 45-50. https://doi.org/10.1016/S0045-6535(98)00166-0
