---
title: "Introduction to eemanalyzeR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to eemanalyzeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`eemanalyzeR` is an R package used providing tools for processing and
analyzing raw excitation-emission matrices (EEMs) and absorbance data.
Includes functions for:

-   data cleaning
-   quality assurance
-   calculation of indices
-   generating visualizations
-   exported cleaned data

The package is designed to creates a streamlined and automated workflow
that ensures consistent analysis and simplifies the coding required to
process the data.

------------------------------------------------------------------------

# The Need

While many instruments to measure EEMs and absorbance come with built in
software to perform post-processing steps, using those tools can lead to
reproducibility issues as the processing steps are often not well
tracked or easily repeatable.

The [drEEM toolbox](https://dreem.openfluor.org/) in MATLAB was one of
the first tools to address this issue, allowing for users to create a
processing script that allowed for replication and consistent methods.
However, MATLAB is a proprietary software, limiting its accessibly.

Two R packages have been created to provide open access processing
tools:

-   [`eemR`](https://CRAN.R-project.org/package=eemR)
-   [`staRdom`](https://CRAN.R-project.org/package=staRdom)

However, these tools require a fair bit of coding knowledge to set up a
processing procedure. The `eemanalyzeR` package aims to be simple enough
that undergraduates, who are often the ones running the instruments, can
process the data to shorten the time between data collection and data
checks to ensure samples can be rerun within hold times.

**This package aims to fill the following gaps:**

-   More comprehensive dealing with absorbance data
-   Applying QA/QC checks to ensure quality data
-   Automate processing via a single function to make it easy for
    beginner coders to use
-   Retain the ability to customize processing steps

------------------------------------------------------------------------

# Installation

Install and load the most recent approved version from CRAN by running

```{r, include=TRUE, eval=FALSE}
install.packages("eemanalyzeR")
```

Install and load the most recent version of `eemanalyzeR` from GitHub by
running

```{r, include=TRUE, eval=FALSE}
# Installing from GitHub requires you first install the remotes package
install.packages("remotes")

# install the most recent version from GitHub
remotes::install_github("katiewampler/eemanalyzeR", ref = "master")
```

Install and load the most recent **development** version of
`eemanalyzeR` from GitHub by running

```{r, include=TRUE, eval=FALSE}
# Installing from GitHub requires you first install the remotes package
install.packages("remotes")

# install the most recent development version from GitHub
remotes::install_github("katiewampler/eemanalyzeR", ref = "dev")
```

Load the package

```{r, include=TRUE, eval=TRUE}
library(eemanalyzeR)
```

------------------------------------------------------------------------

# Processing Data

To go from raw data to processed and clean data use the
[run_eems](../html/run_eems.html) function. At a minimum this requires
the following arguments:

-   **input_dir:** the file path to the folder containing the raw EEMs
    and absorbance files on your computer.
-   **filename:** the name of your project, used to name your exported
    files.

```{r eval=FALSE, message=FALSE, warning=FALSE}
  run_eems(input_dir = system.file("extdata", package = "eemanalyzeR"),
           filename = "eemanalyzeR-example")
```

This function will export the processed data, indices calculations, and
plots by defaults we can look at the function outputs.

```{r eval=FALSE, echo=FALSE}
tree_output <- capture.output(fs::dir_tree(file.path(tempdir(), "eemanalyzeR-example")))
tree_output[1] <- "~/"

tree_output <- tree_output[order(tools::file_ext(tree_output))]
tree_output <- gsub("└──", "├──", tree_output)
tree_output[length(tree_output)] <- gsub("├──", "└──", tree_output[length(tree_output)] )
cat(tree_output, sep = "\n")
```

## The Outputs

-   **\*\_indices_filename.csv**:The absorbance and fluorescence indices
-   **absorbance_plot_filename.png:** Plot showing the absorbance
    spectra
-   **\* .png:** Plots of the individual EEMs
-   **summary_plots_filename.png:** Plot showing all of the EEMs
    together
-   **readme_filename.txt:** A text file detailing the processing steps
    and any warnings that occurred during data processing
-   **processed_data_filename.rds:** An R readable object containing all
    of the exported data

------------------------------------------------------------------------

# Setting up Processing Defaults

`eemanalyzeR` will process files directly "out of the box", but specific
research projects may have differing processing needs. If you would like
to consistently modify the defaults used in the
[run_eems](../html/run_eems.html) function you can use the
[edit_user_config](../html/user_config.html) function. This will create
a YAML file with your selected defaults which is stored on your computer
will be read when you load the package.

------------------------------------------------------------------------

# Creating QA/QC Standards

One of the goals with the `eemanalyzeR` package was to build in QA/QC
checks to ensure data quality. While not required to use the package,
you can use the [create_mdl](../html/create_mdl.html) and
[create_std](../html/create_std.html) functions. These function requires
a directory of analytical blanks and check standards (for instance the
[tea standard](<doi:10.3133/ofr20181096>)
recommended by the USGS). While you can create these standards with any
number of samples, its recommended to use at least 20 for stability.

Use the following code to create the method detection limit (MDL) files.
These by default will be stored in your user-specific data directory and
loaded when needed to process samples.

```{r, eval=TRUE}
#specify where the long-term blank files live
dir <- file.path(system.file("extdata", package = "eemanalyzeR"), "long-term-blanks")

#generate mdl file
eem_mdl <- create_mdl(dir, type = "eem", qaqc_dir = tempdir()) 
abs_mdl <- create_mdl(dir, type = "abs", qaqc_dir = tempdir())
```

We can use a similar process to create the check standard files.

```{r, eval=TRUE}
#specify where the long-term blank files live
dir <- file.path(system.file("extdata", package = "eemanalyzeR"), "long-term-std")

#generate mdl file
eem_std <- create_std(dir, type = "eem", abs_pattern = "ABS", qaqc_dir = tempdir()) 
abs_std <- create_std(dir, type = "abs", abs_pattern = "ABS", qaqc_dir = tempdir())
```

These files will be used when calculating absorbance and fluorescence
indices to ensure:

1.  Sample index values are "real" and distinguishable above a blank
    (above MDL)
    -   Prevents reporting of values that are just noise
2.  The analytical blanks in the sample run are blank (below MDL)
    -   Checks for potential contamination
3.  Check standard index values in the sample run are consistent with a
    long-term standard (within 20 % of mean standard)
    -   Checks for potential run issues (air bubbles, equipment issues,
        calculation issues)

```{r, include=FALSE, eval=TRUE}
#clean up temp files from vignette 

```

