---
title: "Interpreting eemanalyzeR Outputs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpreting eemanalyzeR Outputs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Exported Data

Every data export from `eemanalyzeR` using the `export_data()` or
`run_eems()` functions contains the following files:

-   **\* .png:** Plots of the individual EEMs

    ![Contour plot of a single
    EEM.](eemanalyzeR-example/B1S2ExampleTeaStdSEM.png){width="500"}

-   **absorbance_plot_filename.png:** Plot showing the absorbance
    spectra

    ![Line plot of absorbance for each
    sample.](eemanalyzeR-example/absorbance_plot_eemanalyzeR-example.png){width="500"}

-   **summary_plots_filename.png:** Plot showing all of the EEMs
    together

    ![Four contour plots for each EEM within the example
    dataset.](eemanalyzeR-example/summary_plots_eemanalyzeR-example.png){width="1000"}

-   **\*\_indices_filename.csv**:The absorbance and fluorescence indices

    ```{r echo=FALSE}
    indices <- read.csv("eemanalyzeR-example/fluorindices_eemanalyzeR-example.csv")
    head(indices)
    ```

-   **readme_filename.txt:** A text file detailing the processing steps
    and any warnings that occurred during data processing

    ```{r echo=FALSE}
    readme <- readLines("eemanalyzeR-example/readme_eemanalyzeR-example.txt")
    head(readme)
    ```

-   **processed_data_filename.rds:** An R readable object containing all
    of the exported data. Read in using `readRDS()` function.

    ```{r echo=FALSE}
    data_file <- readRDS("eemanalyzeR-example/processed_data_eemanalyzeR-example.rds")

    if (requireNamespace("lobstr", quietly = TRUE)) {
       lobstr::tree(data_file, max_depth = 1)
    }

    ```

<br>

# QA/QC Checks

## Visual Validation of Instrument Blank

The instrument blank can be plotted and visually validated by the user.
This step is used to visually look for blank contamination or errant
peaks that could impact data during blank subtraction.

-   If this step is performed the **readme** will contain a line that
    this visual check occurred and the user processing the data accepted
    the data.

-   The `validate_blanks()` function also has the option that if the
    user does **not** accept the instrument blank, it may be replaced
    with one of the analytical blanks. In this case the **readme** will
    note the blank that was used instead of the instrument blank.

### What a Blank Should NOT Look Like

#### Broad Blank Contamination

There's a clear peak in the 300-400 nm emission range. This indicates
likely contamination in the blank and it should not be used for blank
subtraction.

![A contour plot with a clear peak in the area where peaks are typically
observed.](figures/blk-cont.png){width="500"}

#### Narrow Blank Contamination

Around excitation 350 nm and emission 500 nm, there's black dot that
sticks out. This is often caused by small particles in the sample
causing small localized areas of high fluorescence. This blank should
also not be used for blank subtraction.

![Contour plot showing a small localized peak at ex: 350 nm, em: 500
nm.](figures/point-blk-cont.png){width="500"}

#### Weird Artifacts

Around emission 420 nm, there's a long "peak" and the intensity is fully
negative. Both these issues indicate likely problems with the instrument
blank and it should not be used for blank subtraction.

![Contour plot showing a strong fluorescence line with negative
intensity.](figures/blk-artifact.png){width="500"}

#### What a Blank Should Look Like

When you remove the scattering lines, the blank should look similar to
the plot below. You're likely to see some random peaks due to noise, but
the noise shouldn't resemble a peak. You might also see strong bars on
the sides of the removed scattering regions. These are normal and just
indicate that the scattering regions were perhaps wider than standard,
but not cause to reject the blank.

![Contour plot showing random noise at low
intensities.](figures/good-blk.png){width="500"}

<br>

## Method Detection Limits

Absorbance and fluorescence indices can be compared to method detection
limits (MDL). The MDL is the minimum signal at which we can confidently
distinguish a measurement from zero and from analytical blanks. Here,
the MDL is calculated using the approach proposed by Hansen et al.
(2018):

$$ \text{MDL} = \text{mean}(\text{long-term blank}) + 3 × \text{SD}(\text{long-term blank})$$
where **long-term blank** is the fluorescence for a specific wavelength
pair or absorbance at a specific wavelength, **mean** is the average
signal, and **SD** is the standard deviation.

To use the MDL checks, the user must create MDL files for both
absorbance and fluorescence data using the `create_mdl()` function. If
no MDL files are found, a warning will appear in the **readme** file.

<br>

## Check Standards

Samples of sample type `check` can also be checked against a long-term
average for consistency.

To use the check standard checks, the user must create average standard
files for both absorbance and fluorescence data using the `create_std()`
function.

-   **If no standard files are found**: a warning will appear in the
    **readme** file.

-   **If a standard file exists**: all available index values will be
    compared between the check standards within the run and the long
    term standard. If the value is outside the tolerance range (default
    is 20%), it will be flagged in the index file. Additionally the
    **readme** file will contain a summary of the number of indices
    outside the tolerance range across the entire run and the tolerance
    used.

<br>

## Index QA/QC Flags

Flags are used to indicate missing or potentially problematic indices.
If the the issue is such that there's no value to report, only the flag
will be returned (i.e., `DOC01`). If there's a value to report, but it
should be used with knowledge there may be an issue with the value, the
flag will append the value (i.e., `0.412_VAL01`) if the indices are in
"wide" format (the default) or in a `QAQC_flag` column in a "long"
format. See `get-indices()` for more details.

**The following flags may appear in the indices output files:**

-   **DATA01**: Missing data required for calculation
-   **DATA02**: Missing required wavelengths; value may be inaccurate
-   **DATA03**: Ratio denominator was zero
-   **DATA04**: Spectral slope could not be calculated
-   **DOC01**: Missing dissolved organic carbon (DOC) data
-   **INF01**: Infinite value
-   **MDL01**: All values below MDL
-   **MDL02**: One or more values below MDL; use cautiously
-   **MDL03**: Ratio index where numerator or denominator was entirely
    below MDL
-   **NEG01**: Negative value
-   **STD01**: Check standard value outside tolerance
-   **VAL01**: Value below typical range
-   **VAL02**: Value above typical range

<br>

# Indices

The absorbance indices reported by the default index method
(`eemanalyzeR`) and their **typical** interpretations are:

```{r echo=FALSE, include=TRUE}
abs_index <- data.frame('Index Name'=c("SUVA 254", "SUVA 280", "SVA 412",
                               "Spectral Slope (275-295)",
                               "Spectral Slope (350-400)",
                               "Spectral Slope Ratio", 
                               "E~2~/E~3~", "E~4~/E~6~"
                               ), 
                        'Index Abbreviation' =c("SUVA254", "SUVA280", "SVA412", "S275_295",
                                    "S350_400", "SR", "E2_E3", "E4_E6"),
                        'Interpretation'=c("proxy for aromaticity", 
                                    "proxy for aromaticity",
                                    "proxy for aromaticity",
                                    "related to molecular weight and aromaticity",
                                    "related to molecular weight and aromaticity", 
                                    "related to molecular weight", 
                                    "related to molecular weight and aromaticity",
                                    "related to humic-like organic matter"
                                    ))
colnames(abs_index) <- c("Index Name", "Index Abbreviation", "Interpretation")
knitr::kable(abs_index)
```

<br>

The fluorescence indices reported by the default index method
(`eemanalyzeR`) and their **typical** interpretations are:

```{r echo=FALSE, include=TRUE}
eem_index <- data.frame('Index Name'=c("Peak B", "Peak T", "Peak A", "Peak M", "Peak C",
                                       "Peak D", "Peak E", "Peak N", 
                                       "Ratio of Peak A to Peak T",
                                       "Ratio of Peak C to Peak A",
                                       "Ratio of Peak C to Peak M",
                                       "Ratio of Peak C to Peak T",
                                       "Fluorescence Index", "Humification Index-Zsolnay",
                                       "Humification Index-Ohno",
                                       "Freshness Index", "Biological Index"), 
                        'Index Abbreviation' =c("pB", "pT", "pA", "pM","pC","pD",
                                                "pE", "pN","rAT", "rCA", "rCM", "rCT", 
                                                "FI", "HIX", "HIX_ohno", "fresh", "BIX"),
                        'Interpretation'=c("tyrosine-like, protein-like organic matter",
                                           "tryptophan-like, protein-like organic matter",
                                           "UV humic-like organic matter",
                                           "marine humic-like organic matter", 
                                           "visible humic-like organic matter", 
                                           "soil fulvic acid-like organic matter",
                                          "soil fulvic acid-like organic matter",
                                          "related to phytoplankton productivity",
                                          "ratio of humic-like to fresh organic matter",
                                          "ratio of humic-like to fulvic-like organic matter",
                                          "amount of blueshifted organic matter",
                                          "related to biochemical oxygen demand",
                                          "terrestrial versus microbial sources", 
                                          "indication of humic substances",
                                          "indication of humic substances", 
                                          "indication of recently produced organic matter",
                                          "indicator of autotrophic activity"

                                    ))
colnames(eem_index) <- c("Index Name", "Index Abbreviation", "Interpretation")
knitr::kable(eem_index)
```

<br>

For more details on interpretation and sources see the vignette
[`eemanalyzeR-indices`](eemanalyzeR-indices.html).

<br>

# References

Hansen, A. M., Fleck, J., Kraus, T. E. C., Downing, B. D., von
Dessonneck, T., & Bergamaschi, B. (2018). *Procedures for using the
Horiba Scientific Aqualog® fluorometer to measure absorbance and
fluorescence from dissolved organic matter* (USGS Numbered Series No.
2018-1096). U.S. Geological Survey.
<https://doi.org/10.3133/ofr20181096>
